---
title: "Argument Checking for Pedagogy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Argument Checking for Pedagogy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(useclinicutils)
```

# Motivation

Bottom-line: *Argument checks in code for learning IS ABSOLUTELY ESSENTIAL*.

```{r motivation}
helloworld <- function(name = "World") sprintf("Hello, %s!", name)
helloworld()
helloworld("Carl")
helloworld(c("Alice", "Bob")) # hmm, maybe not what a student expects ...
try(helloworld(Carl)) # Error in sprintf, not helloworld
```

The `helloworld` task is a ubiquitous starting point when developing
understanding for most programming languages. Yet even in this most introductory
example lurks a potentially inscrutable error for novices.

# Your Package

Many tasks and activities entail some amount of programming, without explicitly
being about programming. That's the sort of area we imagine you're writing a
package about to provide learning activities and/or support for such activities.

Development of the ability to troubleshoot the explicitly programming-related
problems of the kind in the motivating example might be one of your course's
learning objectives. However, often that is *not* an objective, particularly in
courses that are about a topic which _uses_ programming rather than is _about_
programming.

In that _use of_ context, error triage can be an undesirable time sink for
instructors and students. Even for the _about_ context, when working with
beginners or for particularly arcane errors, you might want to provide gentler
alternatives.

Put another way: the pedagogical context means working with novices (even if not
_coding_ novices): helpful misuse checking / error messages is a *MUST-have*.

The `useclinicutils` package provides a light-weight approach to checking
function arguments, designed for use in learning packages generated with the
`useclinic` approach.

Your particular course might need more sophisticated error checking. Maybe even
*less* sophisticated checking - that all depends on your learning objectives. If
you think that's true (minding need to balance also not requiring novices to
install a bunch of other packages), look into e.g.
[this write up](https://blog.r-hub.io/2022/03/10/input-checking/) for
alternatives. If using one of those options, you don't need `checker` and might
not need `useclinicutils`.

# A Light Weight Solution: `checker`s

```{r motivation2}
check_string <- checker(
  is.character(x) && length(x) == 1,
  "'%s' must a character vector of length == 1."
)
helloworld <- function(name = "World") sprintf(
  "Hello, %s!", name |> check_string()
)
helloworld()
helloworld("Carl")
try(helloworld(Carl))
```

Using the `checker`-family of functions, you can quickly define constraint-
checking functions and error messages in your package, then incorporate them in
your exported (for student consumption) code. You can then encourage students to
inspect the functions directly in addition to reading their documentation:

```{r motivation3}
helloworld
```

The constraint checks clarify and reinforce your documentation of the code,
while also affording you the opportunity to write error messages tailored to
your audience.

To use checkers in your package, you still need to define them. We recommend a
`R/checks.R` file, where you write definitions akin to:

```{r examplecheck}

```

# More Pedagogical Motivation

In `R`, defining functions does not generally entail specifying what "type" or
"class" the inputs are, nor are coders required to formally specify the type of
the value that is returned by the function. There are several downsides to this
approach when it comes to learning on some particular code.

In a typed language, the code itself must include the type information in a
particular format, often with keywords, and the documentation and training
materials for that syntax will be more generally available. Your package code
can really only indicate type with variable names (which might detract from
telegraphing other information about those variables) or documentation. But
there aren't standards for documenting type, nor does the population tool
`roxygen2` provide a standard way to capture then output type information.

#' The `checks.R` script is intended to include `check_...` validators that are
#' used within various topic functions. These functions are not intended to be
#' exported (though can be viewed with `{{{pkgname}}}:::check_...`). They are meant
#' to seen by students if they inspect exported functions (i.e. they should be
#' named informatively to indicate what arguments are being checked how), but
#' their guts ignored. More importantly, they should provide informative errors
#' when the functions that *are* exported are misapplied.
#'
#' In general, `check_...`s are intended to:
#'  - concern one feature (therefore, one logical test) at a time
#'  - have a succinct, discernible response when that test is violated
#'  - be applied via pipe in exported functions
#'
#' Note the [get_piped_call_var()] and [checker()] functions are not exported;
#' as a user of this template, you shouldn't (1) touch them or (2) have to
#' understand their internals. All you should need to do is define new checkers
#' like `check_something <- checker(SOMELOGIC, "Some message about %s")`.
#'
#' Lastly: while [checker()] below provides a convenient way to check single
#' arguments, your argument validations may need to pass more than one argument
#' at a time. Several approaches are plausible: you could defined
#' `check_something(x, y)` by hand, you could pass a list combining multiple
#' values, etc: [checker()] is simply a convenience, not a requirement.
#'
#' Using the example `check_...`s at the bottom of this script, a
#' `helloworld(...)` function might look like:
#'
#' @examples
#' helloworld <- function(name) {
#'   name |> check_character() |> check_scalar() |> check_nonemptychar()
#'   sprintf("Hello, %s!", name)
#' }
#'
#' helloworld
#' helloworld("Carl") # works
#' try(helloworld(1)) # nope
#' a <- 1
#' try(helloworld(a)) # nope
#' a <- "Carl"
#' helloworld(a) # works
#' try(helloworld(c("Alice", "Bob"))) # nope
#' try(helloworld("")) # nope

